---
title: "Reports"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reports}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup-load, echo = FALSE}
library(mlr3fairness)
td = "doc"
unlink(list.files(td, full.names = TRUE), recursive = TRUE)
```

`mlr3fairness` contains several templates that allow for creating reports based on `RMarkdown` files.
The `report_*` functions instantiate a new `.Rmd` file that can be further adapted by the user.

The following reports are currently available in `mlr3fairness`.

| Report             | Description             | Reference             |
| ------------------ | ----------------------- | --------------------- |
| `report_modelcard` | Modelcard for ML models | Mitchell et al., 2018 |
| `report_datasheet` | Datasheet for data sets | Gebru et al., 2018    |
| `report_fairness`  | Fairness Report         | –                     |

**Usage:**


Templates contain a set of pre-defined questions which can be used for reporting as well as
initial graphics. The created `.Rmd` file can then be extended by the
user. It can later be converted into a `html` report using `rmarkdown::render()`.


```{r eval_false_example_for_vignette, eval = FALSE, results = 'hide', message = FALSE}
library(mlr3fairness)
rmdfile = report_datasheet()
rmarkdown::render(rmdfile)
```


```{r build_modelcard_example_for_vignette, echo = FALSE, results = 'hide', message = FALSE}
rmdfile = report_modelcard(paste0(td, "/modelcard"))
rmarkdown::render(rmdfile)
```

## Examples

#### [Example: Model Card](`r gsub(".Rmd", ".html", rmdfile)`)


```{r build_datasheet_example_for_vignette, echo = FALSE, results = 'hide', message = FALSE}
rmdfile = report_datasheet(paste0(td, "/datasheet"))
rmarkdown::render(rmdfile)
```

#### [Example: Data Sheet](`r gsub(".Rmd", ".html", rmdfile)`)




```{r build_fairness_example_for_vignette, echo = FALSE, results = 'hide', message = FALSE, warning = FALSE, error = FALSE}
task = tsk("compas")
learner = lrn("classif.rpart", predict_type = "prob")
rr = resample(task, learner, rsmp("cv", folds = 5))
rmdfile = report_fairness(paste0(td, "/fairness"), list(task = task, resample_result = rr))
rmarkdown::render(rmdfile)
```

#### [Example: Fairness Report](`r gsub(".Rmd", ".html", rmdfile)`)
