---
title: "Reports"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{reports}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup-load}
library(mlr3fairness)
```


| Report             | Description             | Reference             |
| ------------------ | ----------------------- | --------------------- |
| `report_modelcard` | Modelcard for ML models | Mitchell et al., 2018 |
| `report_datasheet` | Datasheet for data sets | Gebru et al., 2018    |
| `report_fairness`  | Fairness Report         | –                     |

**Usage:**

The `report_*` functions instantiate a new `.Rmd` template that contains
a set of pre-definedquestions which can be used for reporting as well as
initial graphics. The created `.Rmd` file can then be extended by the
user. It can later be converted into a `html` report using`rmarkdown`’s
`render`.


```{r, eval = FALSE}
rmdfile = report_datasheet()
rmarkdown::render(rmdfile)
```

# Example: Model Card

```{r}
rmdfile = report_modelcard()
rmarkdown::render(rmdfile)
```

# Example: Datasheet

```{r}
rmdfile = report_datasheet()
rmarkdown::render(rmdfile)
```

# Example: Fairness Report


```{r}
rmdfile = report_fairness()
rmarkdown::render(rmdfile)
```
